Extending Camellia for Distributed Stiffness Matrix Determination in DPG

### 2011-12-08 docs/HPC_report/Extending_Camellia.tex
finished optimal test function discussion
git-svn-id: https://svn.ices.utexas.edu/repos/camellia@97 36274a68-8941-4c93-80ef-affaed5139da
Added more discussion of extending Camellia, including some design limitations.
git-svn-id: https://svn.ices.utexas.edu/repos/camellia@98 36274a68-8941-4c93-80ef-affaed5139da

\section{Introduction to DPG}

The Discontinuous Petrov-Galerkin (DPG) method is a finite element method introduced by Demkowicz and Gopalakrishnan in \cite{DPG1} and \cite{DPG2}. The DPG method was first applied to solve the convection-dominated diffusion equation
\[
\div \left(\beta u - \epsilon\grad u\right) = f,
\]
a prototype for shock and boundary layer problem when $\epsilon \ll 1$, using higher order adaptive meshes in \cite{DPG2} and \cite{DPG3}. For standard finite element methods, the convection-dominated diffusion problem suffers from large oscillations that grow in amplitude with $\frac{1}{\epsilon}$. This behavior illustrates an example of the \emph{instability} of standard finite element methods when it comes to problems with small parameters.

DPG is a \emph{discontinuous} Galerkin (DG) method, which means that the solution is allowed to be discontinuous across inter-element boundaries, with continuity weakly enforced through numerical fluxes defined on element boundaries. DPG is also a \emph{Petrov}-Galerkin method, a finite element method in which the test and trial spaces are allowed to differ. Petrov-Galerkin methods allow selection of test functions that improve the behavior of the finite element. The Streamline Upwind Petrov-Galerkin method of Hughes and Brooks \cite{SUPG} is an example of such a method; it aims to eliminate the unstable oscillations in the convection-dominated diffusion problem for Continuous Galerkin (CG) methods.

@@

\section{Introduction to DPG}\label{sec:DPGIntro}
\subsection{Optimal Test Functions}

The definition of an optimal test function is rigorously defined in \cite{DPG2}, and has been well-known in the finite element community for some time. However, this idea proved to be impractical, as the cost to computing a single optimal test function was equivalent to the cost of solving the original problem over the entire mesh. The key contribution of DPG has been to note that the global nature of this problem is due to the continuous nature of the trial space. For a Discontinuous Galerkin method, the test space is also necessarily discontinuous, which decouples the global problem into a local problem over each element.

With respect to optimal test functions, a discontinuous test space allows the test functions to be computed \emph{locally}, on an per-element basis. Not only does this decrease the problem size dramatically, it is also intrinsically parallel - the optimal test functions over each element can be computed completely independently of each other.

With DPG, given any basis $\{u_i\}$ for the trial space, we can compute the $i$th optimal test function over an element by solving the auxiliary problem 
\[
(v_i, \delta v)_V = b(u_i,\delta v), \quad \forall \delta v\in V
\]
where $V$ is the space of admissible test functions (with proper regularity) on the element. For a trial function whose support spans multiple elements, the optimal test function is solved for independently over each element, then set as the union of the test functions over each element the trial function spans. In practice, we solve this problem approximately by taking $V$ to be the space of admissible test functions of order $p+\Delta p$, where $p$ is the order of the trial space. This results in a fairly small local system

@@

\subsection{Camellia/Trilinos implementation}

In this section, we describe the changes that we implemented within Camellia to allow the work to be partitioned and executed efficiently within an MPI environment.

@@

To facilitate experiments with multiple partitioning strategies (and to maintain separation of concerns), we designed an abstract MeshPartitionPolicy class which, given a Mesh and the number of partitions desired, returns a partition of the elements in the mesh.  The Mesh then uses this to generate a partition of the degree-of-freedom coefficients, which induces a partition of the rows in the stiffness matrix.  The latter partition is used in conjunction with an Epetra FECrsMatrix, a distributed stiffness matrix assembly facility within Trilinos.  The FECrsMatrix allows storage of data not owned by the current MPI node; this data is communicated to the owning MPI node during global assembly of the stiffness matrix.  A convenient consequence of this is that we can separate the cost of global assembly from the cost of local stiffness matrix computations; the latter are independent across MPI nodes, and therefore we expect to see perfect scaling for these.  The scaling of the global assembly will depend on the quality of the partition and the cost of communication between MPI nodes.  We do not hope for perfect scaling here, but we do hope to see a cost per dof per node that does not grow too quickly in the total number of nodes.

>>>>

\section{Introduction}

The Discontinuous Petrov-Galerkin (DPG) methodology is a class of finite element methods introduced by Demkowicz and Gopalakrishnan in \cite{DPG1} and \cite{DPG2}. DPG was first applied to solve the convection-dominated diffusion equation
\[
\div \left(\beta u - \epsilon\grad u\right) = f,
\]
a prototype for shock and boundary layer problem when $\epsilon \ll 1$, using higher order adaptive meshes in \cite{DPG2} and \cite{DPG3}. For standard finite element methods, the convection-dominated diffusion problem suffers from large oscillations that grow in amplitude with $\frac{1}{\epsilon}$. This behavior illustrates an example of the \emph{instability} of standard finite element methods when it comes to problems with small parameters.

DPG is a \emph{discontinuous} Galerkin (DG) method, which means that the solution is allowed to be discontinuous across inter-element boundaries, with continuity weakly enforced through numerical fluxes defined on element boundaries. DPG is also a \emph{Petrov}-Galerkin method, a finite element method in which the test and trial spaces are allowed to differ. Petrov-Galerkin methods allow selection of test functions that improve the behavior of the finite element. The Streamline Upwind Petrov-Galerkin method of Hughes and Brooks \cite{SUPG} is an example of such a method; it aims to eliminate the unstable oscillations in the convection-dominated diffusion problem for Continuous Galerkin (CG) methods.  DPG computes test functions on the fly which are\emph{optimal} in that they deliver the best available approximation to the solution, as measured in an energy norm defined by the problem and a test space which may be chosen by the analyst.

@@

\section{DPG: Some Technical Details}\label{sec:DPGIntro}
\subsection{Optimal Test Functions}

The concept of an optimal test function is rigorously defined in \cite{DPG2}, and has been well-known in the finite element community for some time. However, this idea proved to be impractical, as the cost of computing a single optimal test function was equivalent to the cost of solving the original problem over the entire mesh. The key contribution of DPG has been to note that the global nature of this problem is due to the continuous nature of the trial space. For a Discontinuous Galerkin method, the test space is also necessarily discontinuous, which decouples the global problem into a local problem over each element.

With respect to optimal test functions, a discontinuous test space allows the test functions to be computed \emph{locally}, on an per-element basis. Not only does this decrease the problem size dramatically, it is also intrinsically parallel---the optimal test functions for each element can be computed completely independently of each other.

With DPG, given any basis $\{u_i\}$ for the trial space, we can compute the $i$th optimal test function over an element by solving the auxiliary problem 
\[
(v_i, \delta v)_V = b(u_i,\delta v), \quad \forall \delta v\in V
\]
where $V$ is the space of admissible test functions (with proper regularity) on the element. For a trial function whose support spans multiple elements, the optimal test function is solved for independently over each element, then set as the union of the test functions over each element the trial function spans. In practice, we solve this problem approximately by taking $V$ to be the space of admissible test functions of order $p+\Delta p$, where $p$ is the order of the trial space. This results in a small element-local system which is solved to determine a test function corresponding to each degree of freedom on the element. $\Delta p$ is taken to be 2 in this report.

@@

\subsection{Extending Camellia}

In this section, we describe the changes that we implemented within Camellia to allow the stiffness matrix determination to be executed efficiently within an MPI environment.

@@

To facilitate experiments with multiple partitioning strategies (and to maintain separation of concerns), we designed an abstract MeshPartitionPolicy class which, given a Mesh and the number of partitions desired, returns a partition of the elements in the mesh.  The Mesh then uses this to generate a partition of the degree-of-freedom coefficients, which induces a partition of the rows in the stiffness matrix.  The latter partition is used in conjunction with an Epetra FECrsMatrix, a distributed stiffness matrix assembly facility within Trilinos.  The FECrsMatrix allows storage of data not owned by the current MPI node; this data is communicated to the owning MPI node during global assembly of the stiffness matrix.  A convenient consequence of this is that we can separate the cost of global assembly from the cost of local stiffness matrix computations; the latter are independent across MPI nodes, and therefore we expect to see perfect scaling for these.  The scaling of the global assembly will depend on the quality of the partition and the cost of communication between MPI nodes.  We do not hope for perfect scaling here, but we do hope to see a cost per degree of freedom per node that does not grow too quickly in the total number of nodes.

The MeshPartitionPolicy partitions the mesh into sets of elements assigned to an MPI node; for the interface to Epetra, we need a partition of rows of the global stiffness matrix (each of which corresponds to a degree of freedom).  Degrees of freedom corresponding to numerical fluxes are shared by the elements along whose boundaries they lie.  We adopt a simple greedy policy for this---shared d.o.f.s are assigned to the element with the smallest global identifier.  This means that even if the element partitioning were perfectly balanced, some MPI nodes might still have more d.o.f.s assigned to them than others.  This will not affect the amount of work involved in local stiffness matrix computations, but it may affect MPI communication costs during global assembly.

\subsubsection{Design Limitations: Where Amdahl's Law Might Come to Haunt Us}
Here, we note a few features of the present design that may become bottlenecks as we scale up to larger problems and larger numbers of MPI nodes.

At present, our global solve is implemented using the KLU implementation provided by the Amesos package within Trilinos.  This is a direct solve, executed in serial.  Amesos gathers the global stiffness matrix to rank 0, solves, and then scatters the solution according to the partition we have defined.  For our serial computations prior to the present work, the cost of the global solve has been dominated by the cost of the local optimal test function determination; however, now that the latter cost has been distributed using MPI and the former has not, we expect the global solve to become a bottleneck, both in terms of overall execution time and in terms of the sizes of problems that can be solved, since the KLU solve must fit into a single node's memory to be practical.

Similarly, at present the entire mesh is determined and stored on every MPI node.  This has the advantage that no MPI communication is required during mesh construction, but for very large meshes it may become impractical, because the time cost of mesh construction may grow relative to the distributed parts of the computation, and perhaps (for \emph{extremely} large meshes) because the memory required for the mesh may grow too large for storage on a node.

Finally, certain features of the original code depend upon all the solution coefficients being available to the current processor, so for the time being we distribute the entire solution to every MPI node, after KLU has scattered according to the original partitioning.  We hope to eliminate this dependence on having all solution coefficients available, allowing the solution to be distributed according to the same mapping as was used for the stiffness matrix.
